Finiteness in Cubical Type Theory

### 2020-06-29 sections/finiteness-predicates.tex
Update finiteness-predicates.tex

Before we get to those interesting proofs, though, there are some other things
that need defining.
\AgdaDatatype{Fin}, for instance: \(\AgdaDatatype{Fin}\;n\) is the type of
natural numbers smaller than \(n\).
Its definition relies on a type for disjoint union,
\(\AgdaDatatype{\ensuremath{\uplus}}\).
\begin{multicols}{2}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Data/Fin/Base.tex]{fin-def}
  \end{agdalisting} \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Introduction.tex]{disj-union}
  \end{agdalisting}
\end{multicols}

% It is also expressible with only \(\Sigma\):
% \begin{agdalisting}
%   \ExecuteMetaData[agda/Snippets/Introduction.tex]{sigma-disj-union}
% \end{agdalisting}
% Although the inductive type definition is slightly more ergonomic.

After that interlude, we can get back to defining containers.
\begin{definition}[Containers] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a pair
  \(S , P\) where \(S\) is a type, the elements of which are called
  the \emph{shapes} of the container, and \(P\) is a type family on \(S\), where
  the elements of \(P(s)\) are called the \emph{positions} of a container.
  We ``interpret'' a container into a functor defined like so:
  \begin{agdalisting} \label{container-interp}
    \ExecuteMetaData[agda/Container.tex]{container-interp}
  \end{agdalisting}
\end{definition}

The definition of container is a little abstract: it is instructive to think of
it more concretely for the case of lists.
The container representing finite lists is a pair of a natural number \(n\)
representing the length of the list, and a function \(\AgdaDatatype{Fin}\;n\to
A\), representing the indexing function into the list.

\begin{definition}[Container Membership] \label{container-membership}
  Membership of a container can be defined like so:
  \begin{agdalisting} 
    \ExecuteMetaData[agda/Container/Membership.tex]{membership-def}
  \end{agdalisting}
  Where \(x\;\AgdaDatatype{\ensuremath{\in}}\;\mathit{xs}\) is to be read as ``\(x\)
  is in \(\mathit{xs}\)''.
\end{definition}
\begin{definition}[Fibers] \label{fibers}
  A fiber \cite[definition 4.2.4]{hottbook} is defined over some function \(f :
  A \rightarrow B\).
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Introduction.tex]{fiber}
  \end{agdalisting}
\end{definition}

Membership also makes more sense when described concretely in terms of lists.
Understood this way, \(x\;\AgdaDatatype{\ensuremath{\in}}\;\mathit{xs}\) means
``there is an index into \(\mathit{xs}\) such that the index points at an item
equal to \(x\)''.
\subsection{Split Surjections}
We will now look at how this predicate relates to more traditional, classical
notions of finiteness.
In a classical setting we likely wouldn't mention ``lists'' or the like, and
would instead define finiteness based on the existence of some injection or
surjection.
As it turns out, our definition of finiteness here is precisely the same as the
surjection-based one, in quite a deep way!

>>>>

Before we get to those interesting proofs, though, there are some other things
that need defining.
\AgdaDatatype{Fin}, for instance: \(\AgdaDatatype{Fin}\;n\) is the type of
natural numbers smaller than \(n\).
Its definition relies on a type for disjoint union,
\(\AgdaDatatype{\ensuremath{\uplus}}\).
\begin{multicols}{2}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Data/Fin/Base.tex]{fin-def}
  \end{agdalisting} \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Introduction.tex]{disj-union}
  \end{agdalisting}%
  % It is also expressible with only \(\Sigma\):
  % \begin{agdalisting}
  %   \ExecuteMetaData[agda/Snippets/Introduction.tex]{sigma-disj-union}
  % \end{agdalisting}
  % Although the inductive type definition is slightly more ergonomic.
\end{multicols} \noindent
And containers themselves, of course.
Containers are a well-studied topic in dependent type theory, with a rich
theory: we won't dive in to that here.
\begin{definition}[Containers] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a pair
  \(S , P\) where \(S\) is a type, the elements of which are called
  the \emph{shapes} of the container, and \(P\) is a type family on \(S\), where
  the elements of \(P(s)\) are called the \emph{positions} of a container.
  We ``interpret'' a container into a functor defined like so:
  \begin{agdalisting} \label{container-interp}
    \ExecuteMetaData[agda/Container.tex]{container-interp}
  \end{agdalisting}
\end{definition}

The definition of container is a little abstract: it is instructive to think of
it more concretely for the case of lists.
The container representing finite lists is a pair of a natural number \(n\)
representing the length (or ``shape'') of the list, and a function
\(\AgdaDatatype{Fin}\;n\to A\), representing the indexing function into the
list.

One of the nice things about containers is it gives us a generic way to define
``membership'':
\begin{multicols}{2}
  \begin{agdalisting} 
    \ExecuteMetaData[agda/Container/Membership.tex]{membership-def}
  \end{agdalisting} \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Introduction.tex]{fiber}
  \end{agdalisting}
\end{multicols} \noindent 
Here we're using the homotopy-theory notion of a \AgdaDatatype{fiber} to define
membership: a fiber for some function \(f\) and some point \(y\) in its codomain
is a value \(x\) and a proof that \(f \; x \;\AgdaDatatype{≡}\;y\).
Membership also makes more sense when described concretely in terms of lists:
\(x\;\AgdaDatatype{\ensuremath{\in}}\;\mathit{xs}\) means ``there is an index
into \(\mathit{xs}\) such that the index points at an item equal to \(x\)''.
\subsection{Split Surjections}
We will now look at how this predicate relates to more traditional, classical
notions of finiteness.
In a classical setting we likely wouldn't mention ``lists'' or the like, and
would instead define finiteness based on the existence of some injection or
surjection.
As it turns out, our definition of finiteness here is precisely the same as the
surjection-based one, in quite a deep way!
